---
title: "NYPD Shooting Incident Data (Historic)"
author: "T. Stein"
date: "7/13/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=10) 
```

```{r libraries, echo=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(readr)
```

## Step 1 - Import the data
> The first step is to set the url for the NYPD Shooting Incident Data on Data.gov.
```{r url}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
```

> The next step is to import the data.
```{r data}
raw_data <- read_csv(url_in)
```

## Step 2 - Transform and Mutate the Data
> Next we want to select only the columns we will be using for the analysis.
```{r select_data}
filtered_data <- raw_data %>%
  select(-c(INCIDENT_KEY,X_COORD_CD,Y_COORD_CD,Latitude,Longitude,Lon_Lat,LOCATION_DESC))
```

> Next, we want the date column to be of type date.
```{r mutate_date}
filtered_data <- filtered_data %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```  

> Then we will turn categorical variables into factors.
```{r factor_data}
filtered_data <- filtered_data %>%
  mutate(PERP_SEX = as_factor(PERP_SEX)) %>%
  mutate(PERP_RACE = as_factor(PERP_RACE)) %>%
  mutate(PERP_AGE_GROUP = as_factor(PERP_AGE_GROUP)) %>%
  mutate(VIC_AGE_GROUP = as_factor(VIC_AGE_GROUP)) %>%
  mutate(VIC_SEX = as_factor(VIC_SEX)) %>%
  mutate(VIC_RACE = as_factor(VIC_RACE))
```

> After the transformations and mutations have been made, we want to see
a summary of the data.
```{r data_summary}
summary(filtered_data)
```

> Now let's check to see how many ```NA``` values exist in each column.
```{r check_na}
sapply(filtered_data, function(x) sum(is.na(x)))
```

> Then we'll drop the rows with ```NA``` values.
```{r drop_na}
filtered_data <- filtered_data %>%
  drop_na()
```

> Finally we'll check again to see if the ```NA``` values have been dropped.
```{r check_na_again}
sapply(filtered_data, function(x) sum(is.na(x)))
```

> Let's add some extra information about the Boroughs that we'll use in the
analysis.
```{r extra_info}
pop <- c(1438000, 2601000, 1632000, 2299000, 474101)
boro <- c("BRONX", "BROOKLYN", "MANHATTAN", "QUEENS", "STATEN ISLAND")
sqmiles <- c(42.47,69.5,22.82,108.1,58.69)
population <- tibble(population = pop, boro = boro, square_miles = sqmiles)
```

## Step 3 - Visualize the Data
> First, let's plot the number of deaths by police shootings to see if they are 
increasing or decreasing over time by ```BORO```.
```{r plot_deaths_over_time}
filtered_data %>%
  group_by(BORO, YEAR = year(OCCUR_DATE)) %>%
  summarize(DEATHS = sum(STATISTICAL_MURDER_FLAG)) %>%
  ggplot(aes(x=YEAR, y=DEATHS)) +
  geom_point() +
  geom_line() +
  facet_wrap(~BORO) +
  theme_minimal() +
  xlab("Year") +
  ylab("Deaths")
```
*It appears that most boroughs have experienced a decrease in deaths by police 
involved shootings over time. However, in the past year it seems as though all 
of the boroughs have experienced a slight uptick in activity. Brooklyn has 
experienced the most amount of shootings while Staten Island has experienced the 
least amount of shootings.*

>Now we'll plot the number of deaths by ```BORO``` to see which ```BORO``` has the most deaths.
```{r plot_deaths_by_boro}
filtered_data %>%
  group_by(BORO) %>%
  summarize(deaths = sum(STATISTICAL_MURDER_FLAG)) %>%
  inner_join(population, by = c("BORO" = "boro")) %>%
  select(boro = BORO, deaths, population) %>%
  ggplot(aes(x=reorder(boro, -deaths), y=deaths, fill=deaths/population*1000000)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  ggtitle("Number of Deaths by BORO") +
  xlab("Borough") +
  ylab("Murders per Million") +
  guides(fill=guide_legend(title="Deaths Per Million"))
```
*Brooklyn has the most number of deaths by police shooting. Brooklyn also has the
largest population of all the boroughs.*

> We know that Brooklyn has the most deaths by police shooting of all the ```BOROs```. What time of day do most deaths occur?
```{r deadliest_month}
filtered_data %>%
  group_by(hour = hour(OCCUR_TIME)) %>%
  summarize(deaths = sum(STATISTICAL_MURDER_FLAG)) %>%
  ggplot(aes(hour, deaths, fill=deaths)) +
  geom_bar(stat = "identity", show.legend=FALSE) +
  theme_minimal() +
  ggtitle("Deaths by Hour of Day") +
  xlab("Hour of Day") +
  ylab("Deaths")
```
*In Brooklyn, most deaths by police shootings occur between 12:00am - 3:00am and 
5:00pm - 10:00pm.*

## Step 4 - Model the Data
> Is ```BORO``` a good predictor for predicting if you will be killed in a police
involved shooting?
```{r model_boro}
model <- lm(STATISTICAL_MURDER_FLAG ~ BORO, data=filtered_data)
summary(model)
```
*At the 95% confidence interval, ```BORO``` is not statistically significant in 
predicting if you will be killed if involved in a police shooting.*

> Are ```VIC_AGE_GROUP``` and ```VIC_RACE``` good predictors for predicting if 
you will be killed in a police involved shooting?
```{r model_vic}
model <- lm(STATISTICAL_MURDER_FLAG ~ VIC_AGE_GROUP + VIC_RACE, data=filtered_data)
summary(model)
```
*At the 95% confidence interval, ```VIC_AGE_GROUP``` and ````VIC_RACE``` are statistically significant in predicting if you will be killed in a police involved shooting.*

## Step 5 - Identify Bias
> Because I did not pick the topic of NYPD Shooting Incidents, I did not introduce
bias in the selection of the topic or data. I was raised in Manhattan so there 
could be bias introduced in my assumptions of the most dangerous boroughs. The
way I mitigated my bias was by not making assumptions and by coming to conclusions
through proper and ethical data science practices.

## Step 6 - Conclusion
> From the NYPD Shooting Incident data we found some interesting findings. First,
police involved shootings are decreasing over time for all ```BOROs```.Second,
the most police shootings occur in Brooklyn. And the police shootings occur
between 12:00am - 3:00am and 5:00pm - 10:00pm. We found that ```BORO``` alone
is not a statistically significant predictor of death. However, ```VIC_AGE_GROUP``` and ```VIC_RACE``` are statistically significant in predicting if you will be killed in a police involved shooting. The next steps of the analysis would beto find additional information in an attempt to create a more accurate model to help predict the probability of dying in a police involved shooting.